---
title: "Application to BRCA data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Application to BRCA data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE}
library(l1spectralclustering)
library(matrixStats)
library(Hmisc)
library(igraph)
library(FactoMineR)
library(factoextra)
library(limma)
library(fgsea)
library(tidyverse)
```

This vignette details the application of the $\ell_1$-spectral clustering to the BRCA (Breast Cancer) cohort from The Cancer Genome Atlas project (TCGA).

# Data loading and preprocessing

```{r}
data_path <- file.path(path.package("l1spectralclustering"), "data")

# Load expression and ground truth subtype data
load(file.path(data_path, "ProcessedDataBRCA.Rdata")) 
load(file.path(data_path, "RealSubtypes_BRCA.Rdata"))

# Align samples between datasets
Samples <- intersect(names(Subtypes),rownames(MA_TCGA))
Subtypes <- Subtypes[Samples]
MA_TCGA <- MA_TCGA[Samples,]

# Feature selection: Keep top 75% most variable genes
filter <- 75
I <- order(colVars(as.matrix(MA_TCGA)),decreasing = TRUE)[1:(75*ncol(MA_TCGA)/100)]
Data <- MA_TCGA[,I]
```

# Construction of the co-expression network

```{r}
# Correlation matrix
Corr <- rcorr(t(Data))

# Threshold choice
corr_threshold <- seq(0,1,0.05)
pval_threshold <- seq(0,0.05,0.01)

Corr_bis <- Corr$r
I <- which(Corr$P<0.05)
Corr_bis[-I] <- 0

Edges <- c()
for (i in (1:length(corr_threshold))){
  Corr_tmp <- Corr_bis
  
  if (length(which(abs(Corr_tmp) < corr_threshold[i]))!=0){
    Corr_tmp[abs(Corr_tmp) < corr_threshold[i]] <- 0
    Corr_tmp[abs(Corr_tmp) >= corr_threshold[i]] <- 1
    if (length(which(rowSums(Corr_tmp)==0))!=0){
      Corr_tmp <- Corr_tmp[-which(rowSums(Corr_tmp)==0),-which(rowSums(Corr_tmp)==0)]
    }
  } else {
    Corr_tmp <- Corr_tmp
  }
  
  Edges <- c(Edges,length(which(Corr_tmp>0))/2)
}
plot(corr_threshold,Edges,type='l',xlab = "Correlation threshold",ylab="Number of edges")
abline(v=0.8,col="red") # this is the best threshold

Corr_bis[abs(Corr_bis) < 0.80] <- 0
Corr_bis[abs(Corr_bis) >= 0.80] <- 1
diag(Corr_bis) <- 0
```

# Application of the l1-spectral clustering

```{r}
# Prepare the data
patient_names <- rownames(Corr_bis)
A <- Corr_bis
colnames(A) = rownames(A) = NULL

# Run l1-spectral clustering
results <- l1_spectralclustering(A=A, pen="thresholdedLS") 
clusters <- results$comm%*%c(1:ncol(results$comm))
clusters <- as.vector(clusters)
names(clusters) <- patient_names

# Save results
results_path <- file.path(path.package("l1spectralclustering"), "results")
results_file <- file.path(results_path, "RealBRCA_estimatedClusters.Rdata")
save(clusters,file = results_file)
```

# Biological characterization of the subtypes

```{r}
Subtypes <- Subtypes[names(clusters)]

nnclass <- which(clusters==0)
clusters_small <- clusters[-nnclass]
Subtypes_small <- Subtypes[-nnclass]

table(clusters_small,Subtypes_small)
chisq.test(table(clusters_small,Subtypes_small))
```
## Network visualization (Figure 4)

```{r}
Corr_Thresh <- Corr$r
Corr_Thresh[abs(Corr_Thresh) < 0.9] <- 0
Corr_Thresh[abs(Corr_Thresh) >= 0.9] <- 1 # for visualization only
diag(Corr_Thresh) <- 0

Corr_Thresh_small <- Corr_Thresh[names(clusters_small),names(clusters_small)]
I <- which(rowSums(Corr_Thresh_small)==0)
Corr_Thresh_small <- Corr_Thresh_small[-I,-I]
clusters_Thresh_small <- clusters_small[-I]

# 1. Colors 
cluster_colors <- c("#1E3AB1", "#BCF634", "#F64E34", "orange", "purple")

# 2. Create the graph from the matrix
g <- graph_from_adjacency_matrix(as.matrix(Corr_Thresh_small), 
                                 mode = "undirected", 
                                 diag = FALSE)

# 3. Calculate node sizes 
V(g)$size <- (degree(g) / max(degree(g))) * 4 + 1.5

# 4. Assign the colors
V(g)$color <- cluster_colors[as.factor(clusters_Thresh_small)]

# 5. Generate layout
coords <- layout_with_fr(g, niter = 1000)
coords <- coords * 2.5 

# 6. Final plot
pdf(paste0(results_path,"/Patient_Similarity_Network.pdf"), width = 10, height = 10)
plot(g, 
     layout = coords,
     vertex.label = NA,
     vertex.color = V(g)$color,
     vertex.frame.color = "white", 
     vertex.frame.width = 0.3,
     vertex.size = V(g)$size,
     edge.color = adjustcolor("gray40", alpha.f = 0.4), 
     edge.width = 0.6,
     main = "Network")
dev.off()
```

## PCA visualization (Figure 5)

```{r}
# PCA
PCA_df <- as.data.frame(MA_TCGA[names(clusters_small), ])
PCA_df$Clusters <- factor(clusters_small)

res.pca <- PCA(PCA_df, quali.sup = "Clusters", graph = FALSE)

PCA_fig <- fviz_pca_biplot(res.pca, 
                           # Individuals
                           habillage = "Clusters", 
                           addEllipses = TRUE, 
                           ellipse.level = 0.25,
                           geom = "point",
                           pointsize = 2,
                           
                           # Variables (Genes)
                           select.var = list(contrib = 10), 
                           col.var = "black", 
                           repel = TRUE, 
                           labelsize = 6,
                           
                           # Aesthetics
                           title = "PCA Biplot") +
  
  # 1. Custom High-Contrast Palette
  # 1: Royal Blue, 2: Lime Green, 3: Crimson, 4: Vivid Orange, 5: Forest Green
  scale_color_manual(name = "Clusters", 
                     labels = c("1", "2", "3", "4", "5"),
                     values = c("#1E3AB1", "#BCF634", "#F64E34", "orange","purple")) +
  
  # 2. Matching Fill for Ellipses
  scale_fill_manual(name = "Clusters", 
                    labels = c("1", "2", "3", "4", "5"),
                    values = c("#1E3AB1", "#BCF634", "#F64E34", "orange","purple")) +
  
  # 3. Shape assignment (19: circle, 17: triangle, 15: square, 3: plus, 18: diamond)
  scale_shape_manual(values = c(19, 17, 15, 3, 18)) +
  
  # 4. Hide the redundant shape legend
  guides(shape = "none") +
  
  theme_minimal() +
  theme(legend.position = "right") +
  theme(
    axis.title = element_text(size = 20, face = "bold"), # Specific size for Axis Labels
    axis.text = element_text(size = 16),                # Specific size for Axis Numbers
    legend.title = element_text(size = 18),             # Specific size for Legend Title
    legend.text = element_text(size = 16)               # Specific size for Legend Items
  )

PCA_fig

ggsave(filename = c("PCA_figure.pdf"), 
       path = results_path, 
       plot = PCA_fig,
       device = "pdf",
       width = 12,                   
       height = 8, 
       units = "in",
       dpi = 300                     
)
```

## Functional enrichment (GSEA)
```{r}
# Load pathways once
pathway_file <- file.path(data_path, "h.all.v2026.1.Hs.symbols.gmt")
pathways <- gmtPathways(pathway_file)

n_clusters <- length(table(clusters_small))

for (c in (1:n_clusters)){
  # 1. Differential expression (one vs all)
  group <- factor(ifelse(clusters_small == c, "Cluster_c", "Base"))
  design <- model.matrix(~group)

  data <- MA_TCGA
  data <- data[names(clusters_small),]
  fit <- lmFit(t(data), design)
  fit <- eBayes(fit)

  # 2. Rank genes
  stats <- topTable(fit, coef = 2, number = Inf)
  stats$rank_metric <- sign(stats$logFC) * -log10(stats$P.Value)
  stats_sorted <- stats[order(-stats$rank_metric), ]
  ranked_list <- stats_sorted$rank_metric
  names(ranked_list) <- rownames(stats_sorted)

  # 3. Run GSEA
  gsea_results <- fgsea(pathways = pathways, 
                      stats = ranked_list,
                      minSize = 15,
                      maxSize = 500)

  significant_pathways <- gsea_results[gsea_results$padj < 0.05, ]
  significant_pathways <- significant_pathways[order(significant_pathways$NES, decreasing = TRUE), ]

  # 4. Plot the results
  plot_df <- as.data.frame(significant_pathways)

  plot_df$pathway <- gsub("HALLMARK_", "", plot_df$pathway)
  plot_df$pathway <- gsub("_", " ", plot_df$pathway)

  plot_df$Direction <- ifelse(plot_df$NES > 0, "Upregulated", "Downregulated")

  g <- ggplot(plot_df, aes(x = reorder(pathway, NES), y = NES, fill = Direction)) +
    geom_bar(stat = "identity", width = 0.8) +
    coord_flip() + 
    scale_fill_manual(values = c("Upregulated" = "#E41A1C", 
                               "Downregulated" = "#377EB8")) +
  theme_minimal() +
  labs(
    title = paste0("Hallmark pathway enrichment for cluster ",c),
    subtitle = "Significance threshold: padj < 0.05",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  theme(
    axis.text.y = element_text(size = 10, face = "bold"),
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  )
  g
  
  ggsave(filename = paste0("GSEA_cluster",c,".pdf"), 
       path = results_path, 
       plot = g,
       device = "pdf",
       width = 12,                   
       height = 8, 
       units = "in",
       dpi = 300                     
  )
}
```

