---
title: "Application to BRCA data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Application to BRCA data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE}
library(l1spectralclustering)
library(matrixStats)
library(FactoMineR)
library(factoextra)
library(limma)
library(fgsea)
library(tidyverse)
```

This vignette details the application of the $\ell_1$-spectral clustering to the BRCA (Breast Cancer) cohort from The Cancer Genome Atlas project (TCGA).

# Data loading and preprocessing

```{r}
data_path <- file.path(path.package("l1spectralclustering"), "data")

# Load expression and ground truth subtype data
load(file.path(data_path, "ProcessedDataBRCA.Rdata")) 
load(file.path(data_path, "RealSubtypes_BRCA.Rdata"))

# Align samples between datasets
Samples <- intersect(names(Subtypes),rownames(MA_TCGA))
Subtypes <- Subtypes[Samples]
MA_TCGA <- MA_TCGA[Samples,]

# Feature selection: Keep top 75% most variable genes
filter <- 75
I <- order(colVars(as.matrix(MA_TCGA)),decreasing = TRUE)[1:(75*ncol(MA_TCGA)/100)]
Data <- MA_TCGA[,I]
```

# Construction of the co-expression network

```{r}
# Thresholded correlation matrix
Corr <- cor(t(Data))
Corr[abs(Corr) < 0.70] <- 0
Corr[abs(Corr) >= 0.70] <- 1
diag(Corr) <- 0
```

# Application of the l1-spectral clustering

```{r}
# Run l1-spectral clustering
clusters <- l1_spectralclustering(A=Corr, pen="thresholdedLS", stab=FALSE) # to verify
# and add the code to plot the figure

# Save results
results_path <- file.path(path.package("l1spectralclustering"), "results")
results_file <- file.path(results_path, "RealBRCA_estimatedClusters.Rdata")
save(clusters,file = results_file)
```

# Biological characterization of the subtypes

## PCA visualization (Figure 5)

```{r}
# PCA
PCA_df <- as.data.frame(MA_TCGA[names(clusters), ])
PCA_df$Clusters <- factor(clusters)

res.pca <- PCA(PCA_df, quali.sup = "Clusters", graph = FALSE)

PCA_fig <- fviz_pca_biplot(res.pca, 
                           # Individuals
                           habillage = "Clusters", 
                           addEllipses = TRUE, 
                           ellipse.level = 0.5,
                           geom = "point",
                           
                           # Variables (Genes)
                           select.var = list(name = c("MCM2", "MKI67", "ETS1")), 
                           col.var = "black", 
                           repel = TRUE, # Avoid label overlapping
                           
                           # Aesthetics
                           title = "PCA Biplot") +
  scale_color_manual(name = "Clusters", 
                     labels = c("1", "2", "3", "4"),
                     values = c("#1E3AB1", "#BCF634", "#F64E34", "orange")) +
  scale_fill_manual(name = "Clusters", 
                    labels = c("1", "2", "3", "4"),
                    values = c("#1E3AB1", "#BCF634", "#F64E34", "orange")) +
  guides(shape = "none") +
  theme_minimal()

PCA_fig

ggsave(filename = c("PCA_figure.pdf"), 
       path = results_path, 
       plot = PCA_fig,
       device = "pdf",
       width = 12,                   
       height = 8, 
       units = "in",
       dpi = 300                     
)
```

## Functional enrichment (GSEA)
```{r}
# Load pathways once
pathway_file <- file.path(data_path, "h.all.v2026.1.Hs.symbols.gmt")
pathways <- gmtPathways(pathway_file)

n_clusters <- length(table(clusters))

for (c in (1:n_clusters)){
  # 1. Differential expression (one vs all)
  group <- factor(ifelse(clusters == c, "Cluster_c", "Base"))
  design <- model.matrix(~group)

  fit <- lmFit(t(MA_TCGA[names(clusters),]), design)
  fit <- eBayes(fit)

  # 2. Rank genes
  stats <- topTable(fit, coef = 2, number = Inf)
  stats$rank_metric <- sign(stats$logFC) * -log10(stats$P.Value)
  stats_sorted <- stats[order(-stats$rank_metric), ]
  ranked_list <- stats_sorted$rank_metric
  names(ranked_list) <- rownames(stats_sorted)

  # 3. Run GSEA
  gsea_results <- fgsea(pathways = pathways, 
                      stats = ranked_list,
                      minSize = 15,
                      maxSize = 500)

  significant_pathways <- gsea_results[gsea_results$padj < 0.05, ]
  significant_pathways <- significant_pathways[order(significant_pathways$NES, decreasing = TRUE), ]

  # 4. Plot the results
  plot_df <- as.data.frame(significant_pathways)

  plot_df$pathway <- gsub("HALLMARK_", "", plot_df$pathway)
  plot_df$pathway <- gsub("_", " ", plot_df$pathway)

  plot_df$Direction <- ifelse(plot_df$NES > 0, "Upregulated", "Downregulated")

  g <- ggplot(plot_df, aes(x = reorder(pathway, NES), y = NES, fill = Direction)) +
    geom_bar(stat = "identity", width = 0.8) +
    coord_flip() + 
    scale_fill_manual(values = c("Upregulated" = "#E41A1C", 
                               "Downregulated" = "#377EB8")) +
  theme_minimal() +
  labs(
    title = paste0("Hallmark pathway enrichment for cluster ",c),
    subtitle = "Significance threshold: padj < 0.05",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  theme(
    axis.text.y = element_text(size = 10, face = "bold"),
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  )
  g
  
  ggsave(filename = paste0("GSEA_cluster",c,".pdf"), 
       path = results_path, 
       plot = g,
       device = "pdf",
       width = 12,                   
       height = 8, 
       units = "in",
       dpi = 300                     
  )
}
```

