---
title: "Create simulated data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create simulated data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, message=FALSE, warning=FALSE}
library(l1spectralclustering)
```

This vignette details the generation of synthetic networks used for benchmarking. The process involves creating a base community structure and then applying perturbations controlled by internal connectivity ($p_{in}$) and external noise ($p_{out}$).

# Simple simulation example

This code can be used to generate a single dataset. In this example, we set the density $d$ and the number of nodes $n$ to create a graph with a specific number of clusters ($k=n\times d$).

```{r}
# Set the parameters to simulate data
d <- 0.1 # density (n*d = k, number of clusters)
n <- 100 # number of nodes
p <- list(p_inside = 0.1, p_outside = 0.05) # perturbations
rngSEED <- 12345L

# Run the code to simulate the data
data <- CreateDataSet(k = d*n, n, p, rngSEED)
```


# Full combinatorial grid simulation

To robustly evaluate performance, we simulate data across a wide range of parameter combinations. For every set of parameters, we generate $100$ independent perturbed graphs. 

```{r}
# data directory
path_data <- here::here("data")

# define the parameters
density_range <- c(0.005,0.01,0.02,0.05,0.1,0.2) # density of the graph (k/n)
n_range <- c(50,100,500,1000) # number of nodes
p_inside_range <- c(0.01,0.1,0.25,0.5)
p_outside_range <- c(0.01,0.1,0.25,0.5)
rngSEED <- 12345L
n_graphs <- 100 # number of perturbed graphs to simulate

# run the code to simulate the data
for(n in n_range){
  for(d in density_range){
    for(p_inside in p_inside_range){
      for (p_outside in p_outside_range){
        if (d*n <= 2 || d*n > 25) {
          message(sprintf("Skipping: n=%d, d=%.2f (k=%.1f out of bounds)", n, d, d*n))
          next 
        }
        graphs <- list() # to store the simulated graphs
        
        ClustersLength <- rep(floor(1/d),d*n)
        
        if (length(ClustersLength)>1){
          # there should be at least two clusters
          if (sum(ClustersLength)!=n){
            ClustersLength[length(ClustersLength)] <- n-sum(ClustersLength[-length(ClustersLength)])
          }
          data <- CreateDataSet(k=d*n, n=n, p=list(p_inside=p_inside,p_outside=p_outside), print.plot = TRUE, ClustersLength = ClustersLength)
          ClustersLength <- data$ClustersLength
          
          graphs <-  c(graphs,list(data))
          for(c in 2:n_graphs){
            data <- CreateDataSet(k=d*n, n=n, p=list(p_inside=p_inside,p_outside=p_outside), print.plot = FALSE, ClustersLength = ClustersLength)
            
            graphs <-  c(graphs,list(data))
          }
          names(graphs) <- paste0("Graph",1:n_graphs)
          
          # define the specific filename for this (n, density, noise) combination
          file_name <- sprintf("data_n=%d_density=%.2f_p_inside=%.2f_p_outside=%.2f.Rdata", 
                     n, d, p_inside, p_outside)

          # construct the full destination path
          full_path <- file.path(path_data, file_name)

          # save the 100 perturbed graphs
          save(graphs, file = full_path)
        }
      }
    }
  }
}
```