---
title: "Plot results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, message=FALSE, warning=FALSE}
library(l1spectralclustering)
```

This vignette describes the workflow for generating the performance plots found in Figure 2 of the manuscript. The objective is to evaluate the robustness of the $\ell_1$-spectral clustering framework against state-of-the-art methods, including: spectral clustering algorithm (standard, regularized and robust versions), VGAE (Variational Graph Auto-Encoders), MCL (Markov Cluster Algorithm), Spectrum and Hybrid approaches.
Before visualization, ensure your simulation outputs are stored in a structured directory (`/data` for ground truth and `/results` for outputs). The functions expect a naming convention that includes $n$, $d$, and the perturbation probabilities.

The `merge_simulation_results` function crawls the results directory to aggregate the results into a data frame. In addition to standard metrics (mean/median AMI), it computes the missclassification counts (number of nodes incorrectly assigned to a cluster), the recall (the ratio of classified nodes to the total number of nodes) and 
the AMI calculated only on the subset of nodes actually classified by the algorithm.

```{r}
# Define the grid of parameters used in the simulation
p_inside_range = p_outside_range <- c(0.01, 0.1, 0.25, 0.5)
n <- 500
d <- 0.01
path <- file.path(path.package("l1spectralclustering"))
method = c("l1Spectral", "Spectral", "regSpectral","robustSpectral", "VGAE", "ST_l1Spectral", "Spectrum", "Hybrid", "MCL")

# Compile results into a structured data frame
sim_data <- merge_simulation_results(
  path = path,
  n = n,
  d = d,
  p_inside_range = p_inside_range,
  p_outside_range = p_outside_range,
  method = method
)

```