---
title: "Run simulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, message=FALSE, warning=FALSE}
library(l1spectralclustering)
```

The following code demonstrates the workflow for executing a single simulation experiment across multiple graph instances. To run it, the specific simulation parameters (such as the number of nodes $n$, the density of the graph $d$, and perturbation levels $p_{in}$ and $p_{out}$) and select the desired clustering algorithm (e.g., `l1Spectral`) at the beginning of the script.

```{r}
library(foreach)
library(doParallel)

# Define simulation parameters
n <- 500
d <- 0.02
p_inside <- 0.01
p_outside <- 0.01
current_method <- "l1Spectral"

# Construct the filenames and paths
data_path <- file.path(path.package("l1spectralclustering"), "data")
results_path <- file.path(path.package("l1spectralclustering"), "results")
file_suffix <- sprintf("n=%d_density=%.2f_p_inside=%.2f_p_outside=%.2f.Rdata", n, d, p_inside, p_outside)
data_file    <- file.path(data_path, paste0("data_", file_suffix))
results_file <- file.path(results_path, paste0("results_", current_method,"_",file_suffix))

# Load the simulated data
if (file.exists(data_file)) {
  load(data_file)
  data_example <- graphs[[1]]
} else {
  stop("Simulated data file not found at: ", data_file)
}

# Run the code on the length(graphs) graphs
num_cores <- parallel::detectCores() - 2 
cl <- makeCluster(num_cores)
registerDoParallel(cl)

results <- foreach(l = 1:length(graphs), 
                   # Export ALL your custom functions
                   .export = c("RunSimulations", "Spectral", "regSpectral", "ComputePerformance"),
                   # Load ALL necessary libraries on the workers
                   .packages = c("kernlab", "greed", "igraph", "aricode", "purrr")
) %dopar% {
  
  graph_obj <- graphs[[l]]
  res <- RunSimulations(graph_obj, method = current_method)
  
  return(res)
}
names(results) <- paste0("Graphs",1:length(graphs))

stopCluster(cl)

# Extract each metric 
NMI  <- sapply(results, function(x) x$NMI)
ARI  <- sapply(results, function(x) x$ARI)
AMI  <- sapply(results, function(x) x$AMI)
Time <- sapply(results, function(x) x$Time)

results <- c(results, list(NMI = NMI, AMI = AMI, ARI = ARI, Time = Time))

# Save the results
save(results, results, file = results_file)

cat("Saved results to:", results_file, "\n")
```

This code builds upon the first one by looping over defined ranges of $n$, $d$, $p_{in}$, and $p_{out}$ using the self-tuned version of the $\ell_1$-spectral clustering algorithm. This specific configuration was used to generate Table 1.

```{r}
library(foreach)
library(doParallel)

# Define simulation parameters
n_range <- c(50,100,500,1000)
d_range <- c(0.01,0.02,0.1,0.2)
p_inside_range <- c(0.01,0.1,0.25,0.5)
p_outside_range <- c(0.01,0.1,0.25,0.5)
current_method <- "ST_l1Spectral"

for (n in (n_range)){
  for (d in (d_range)){
    for (p_inside in p_inside_range){
      for (p_outside in p_outside_range){
        # Construct the filenames and paths
        data_path <- file.path(path.package("l1spectralclustering"), "data")
        results_path <- file.path(path.package("l1spectralclustering"), "results")
        file_suffix <- sprintf("n=%d_density=%.2f_p_inside=%.2f_p_outside=%.2f.Rdata", n, d, p_inside, p_outside)
        data_file    <- file.path(data_path, paste0("data_", file_suffix))
        results_file <- file.path(results_path, paste0("results_", current_method,"_",file_suffix))
        
        # Load the simulated data
        if (file.exists(data_file)) {
          load(data_file)
          data_example <- graphs[[1]]
        } else {
          stop("Simulated data file not found at: ", data_file)
        }
        
        # Run the code on the length(graphs) graphs
        num_cores <- parallel::detectCores() - 2 
        cl <- makeCluster(num_cores)
        registerDoParallel(cl)
        
        results <- foreach(l = 1:length(graphs), 
                           # Export ALL your custom functions
                           .export = c("RunSimulations", "Spectral", "regSpectral", "ComputePerformance"),
                           # Load ALL necessary libraries on the workers
                           .packages = c("kernlab", "greed", "igraph", "aricode", "purrr")
        ) %dopar% {
          
          graph_obj <- graphs[[l]]
          res <- RunSimulations(graph_obj, method = current_method)
          
          return(res)
        }
        names(results) <- paste0("Graphs",1:length(graphs))
        
        stopCluster(cl)
        
        # Extract each metric 
        NMI  <- sapply(results, function(x) x$NMI)
        ARI  <- sapply(results, function(x) x$ARI)
        AMI  <- sapply(results, function(x) x$AMI)
        Time <- sapply(results, function(x) x$Time)
        
        results <- c(results, list(NMI = NMI, AMI = AMI, ARI = ARI, Time = Time))
        
        # Save the results
        save(results, results, file = results_file)
        
        cat("Saved results to:", results_file, "\n")       
      }
    }
  }
}
```

The final script executes the full benchmark across all algorithms discussed in the paper. This provides the data necessary for the comparative analysis in Table 2 and the performance visualizations in Figure 3.

```{r}
library(foreach)
library(doParallel)

# Define simulation parameters
settings <- list(c(n=100, d=0.1),c(n=500, d=0.02))
p_inside_range <- c(0.01,0.1,0.25,0.5)
p_outside_range <- c(0.01,0.1,0.25,0.5)
methods <- c("l1Spectral", "Spectral", "regSpectral","robustSpectral","ST_l1spectral", "ST_Spectral", "Hybrid", "MCL")

for (s in (settings)){
  for (current_method in (methods)){
    for (p_inside in p_inside_range){
      for (p_outside in p_outside_range){
        n <- s["n"]
        d <- s["d"]
        # Construct the filenames and paths
        data_path <- file.path(path.package("l1spectralclustering"), "data")
        results_path <- file.path(path.package("l1spectralclustering"), "results")
        file_suffix <- sprintf("n=%d_density=%.2f_p_inside=%.2f_p_outside=%.2f.Rdata", n, d, p_inside, p_outside)
        data_file    <- file.path(data_path, paste0("data_", file_suffix))
        results_file <- file.path(results_path, paste0("results_", current_method,"_",file_suffix))
        
        # Load the simulated data
        if (file.exists(data_file)) {
          load(data_file)
          data_example <- graphs[[1]]
        } else {
          stop("Simulated data file not found at: ", data_file)
        }
        
        # Run the code on the length(graphs) graphs
        num_cores <- parallel::detectCores() - 2 
        cl <- makeCluster(num_cores)
        registerDoParallel(cl)
        
        results <- foreach(l = 1:length(graphs), 
                           # Export ALL your custom functions
                           .export = c("RunSimulations", "Spectral", "regSpectral", "ComputePerformance"),
                           # Load ALL necessary libraries on the workers
                           .packages = c("kernlab", "greed", "igraph", "aricode", "purrr")
        ) %dopar% {
          
          graph_obj <- graphs[[l]]
          res <- RunSimulations(graph_obj, method = current_method)
          
          return(res)
        }
        names(results) <- paste0("Graphs",1:length(graphs))
        
        stopCluster(cl)
        
        # Extract each metric 
        NMI  <- sapply(results, function(x) x$NMI)
        ARI  <- sapply(results, function(x) x$ARI)
        AMI  <- sapply(results, function(x) x$AMI)
        Time <- sapply(results, function(x) x$Time)
        
        results <- c(results, list(NMI = NMI, AMI = AMI, ARI = ARI, Time = Time))
        
        # Save the results
        save(results, results, file = results_file)
        
        cat("Saved results to:", results_file, "\n")       
      }
    }
  }
}
```