---
title: "Technical illustration of the penalty selection and stability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Technical illustration of the penalty selection and stability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, message=FALSE, warning=FALSE}
library(l1spectralclustering)
library(glmnet)
library(cvTools)
library(dplyr)
library(scales)
```

This vignette demonstrates the behavior of the $\ell_1$ (Lasso) and thresholded least-squares penalties in the context of the $\ell_1$-spectral clustering. We visualize how the coefficients evolve and how optimal parameters $\lambda$ and $t$ are selected. This code was used to generate Figure 1 of the manuscript.

# Simulation setup

We investigate the stability of the penalty selection under four levels of perturbation ($p \in \{0.01,0.1,0.25,0.5\}$).

```{r}
# Accessing data through the package structure
data_path <- system.file("data", package = "l1spectralclustering")
p_range <- c(0.01, 0.1, 0.25, 0.5)
```

# Main loop

The following code iterates through the perturbation levels to extract the Lasso paths and the thresholded least-squares stability zones.

```{r}
all_data <- list()
optima_list <- list()

for (p in p_range){
  p_str <- sprintf("%.2f", p) 
  
  #file_name <- file.path(data_path, paste0("data_n=500_k=10_p_inside=", p_str, "_p_outside=", p_str, ".Rdata"))
  #load(file_name)
  load(paste0("/Users/mchampion/Documents/GitHub/l1spectralclustering/data/data_n=500_k=10_p_inside=", p_str, "_p_outside=", p_str, ".Rdata"))
  
  A <- graphs$Graph2$A_hat
  n <- ncol(A)
  k <- length(graphs$Graph2$ClustersLength)
    
  Structure <- l1spectral::FindStructure(A)
  clusters <- l1spectral::FindNbrClusters(A, structure  = Structure, k = k)
  Elements <- l1spectral::FindElement(A = A, structure = Structure, clusters = clusters)
    
  index <- Elements$indices$Component1[1]
  index_neighbors <- which(graphs$Graph2$clusters == graphs$Graph2$clusters[index])
    
  svd <- eigen(A)
  eigenvalues <- sort(svd$values,index.return=TRUE)
  eigenvectors <- svd$vectors[,eigenvalues$ix]
    
  U <- t(eigenvectors[,1:(n-k)])
    
  w <- U[,index]
  W <- matrix(U[,-index],ncol=(ncol(U)-1),nrow=nrow(U))
    
  # 1. Lasso solution path
  lasso <- glmnet(-W,w)
  Beta <- lasso$beta
    
  # 2. Cross-validation for Lambda
  lassosol <- cv.glmnet(W,-w)
  cvtop<- min(lassosol$cvm)+5*(max(lassosol$cvm)-min(lassosol$cvm))/100
  error <- min(lassosol$cvm[lassosol$cvm>cvtop])
  lambda_opt <- lassosol$lambda[which(lassosol$cvm==error)]
  I <- which(lasso$lambda==lambda_opt)
  l1_norm_opt <- colSums(Beta)[I]

  # 3. Thresholded least-squares
  groups <- cvFolds(nrow(W),K=5)
  T_grid <- seq(from=0, to=1,0.01)
      
  error_cv <- sapply(T_grid, function(t) {
        sapply(1:5, function(g) {
            train <- which(groups$which != g)
            test <- which(groups$which == g)
            sol <- as.vector(glmnet(W[train,], -w[train], lambda=0)$beta)
            sol[abs(sol) < t] <- 0
            sqrt(mean((W[test,] %*% sol - (-w[test]))^2))
        }) %>% mean()
    })

  t_opt <- T_grid[which.min(error_cv)]
  
  # Collect results for plotting  
  tmp_df <- data.frame(
        L1norm = rep(colSums(abs(Beta)), each = n-1),
        Coeff = as.vector(Beta),
        Number = rep(1:(n-1), ncol(Beta)),
        Perturbation = paste0("p=", p_str)
    ) %>%
    mutate(IsSignal = Number %in% index_neighbors)
    
    all_data[[p_str]] <- tmp_df
    optima_list[[p_str]] <- data.frame(Perturbation = paste0("p=", p_str), 
                                       v_line = l1_norm_opt, h_line = t_opt)
}
plot_data <- bind_rows(all_data)
optima_df <- bind_rows(optima_list)
```

# Visualizing the selection stability

The plot below shows the Lasso paths (lines). Red paths represent the signal (true neighbors), while black paths represent noise. The dashed green lines indicate the optimal $\ell_1$-norm and the threshold $t$ selected by the package.

```{r}
sym_sqrt_trans <- trans_new(
    name = "sym_sqrt",
    transform = function(x) sign(x) * sqrt(abs(x)), 
    inverse = function(x) sign(x) * x^2             
)

tol <- 0.10 # 10% tolerance window
rect_plot_robust <- optima_df %>%
    mutate(
        # Vertical Stability Zone (L1-norm penalty)
        xmin = v_line * (1 - tol),
        xmax = v_line * (1 + tol),
        
        # Horizontal Stability Zone (Coefficient threshold t)
        ymin = h_line * (1 - tol),
        ymax = h_line * (1 + tol)
    )
label_data <- optima_df %>%
    mutate(
        # Move v label to the top right of the crosshair
        v_label_x = v_line * 1.1,
        # Move h label slightly above the crosshair
        h_label_y = h_line + 0.05
    )
ggplot() +
    # Robust Vertical Zone
    geom_rect(data = rect_plot_robust, 
              aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
              fill = "gray80", alpha = 0.4) +
    
    # Robust Horizontal Zone
    geom_rect(data = rect_plot_robust,
              aes(xmin = 0.1, # Use a small number >0 for log1p transform compatibility
                  xmax = max(plot_data$L1norm), 
                  ymin = ymin, ymax = ymax),
              fill = "gray70", alpha = 0.2) +
    
    # The Lasso Paths
    geom_line(data = plot_data, 
              aes(x = L1norm, y = Coeff, group = Number, 
                  color = IsSignal, linewidth = IsSignal, alpha = IsSignal),
              show.legend = FALSE) +
    
    # Decision Crosshairs (Dashed Green) 
    # Vertical (l1_opt)
    geom_vline(data = optima_df, aes(xintercept = v_line), 
               color = "forestgreen", linetype = "dashed", linewidth = 0.8) +
    # Horizontal (t_opt)
    geom_hline(data = optima_df, aes(yintercept = h_line), 
               color = "forestgreen", linetype = "dashed", linewidth = 0.8) +
    
    # Text Labels 
    # Label t_opt (placed on the left side)
    geom_text(data = label_data, 
              aes(x = 0.1, y = h_line, label = "t[opt]"), 
              color = "forestgreen", vjust = -0.5, hjust = 0, parse = TRUE, size = 3.5) +
    # Label l1_opt (placed vertically)
    geom_text(data = label_data, 
              aes(x = v_line, y = 1.0, label = "l[1*','*opt]"), 
              color = "forestgreen", angle = 90, vjust = -0.5, parse = TRUE, size = 3.5) +
    
    # Scales and Aesthetics 
    # Map aesthetics manually: Red=thick/solid, Black=thin/transparent
    scale_linewidth_manual(values = c("FALSE" = 0.3, "TRUE" = 0.3)) +
    scale_alpha_manual(values = c("FALSE" = 0.2, "TRUE" = 1.0)) +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    
    scale_x_continuous(trans = "log1p") + 
    scale_y_continuous(trans = sym_sqrt_trans, 
                       breaks = c(-1, -0.5, 0, 0.5, 1)) +
    
    facet_wrap(~Perturbation, scales = "fixed") + 
    theme_minimal() +
    labs(x = "l1-norm", y = "Coefficient value") +
    theme(panel.grid.minor = element_blank())
```